Solar Flares Across the Spectrum: Observations and Instrumentation

### 2025-04-11 chapter6/chap6.tex
Local changes

%%---------------------------------------------------------------
\begin{equation*}
\Omega_{ij}=\frac{8\pi}{\sqrt{3}}~\frac{I_{H}}{\Delta \epsilon_{ij}}g\omega_{i}~f_{ij}
\end{equation*}
%%---------------------------------------------------------------

\noindent Here, $I_{H}$ denotes the ionization energy of hydrogen, $\Delta \epsilon_{ij}$ represents the threshold energy for the transition, $g$ is the Gaunt factor, $\omega_{i}$ is the statistical weight of the level, and $f_{ij}$ stands for the oscillator strength. In optically thin conditions, the intensity ratio of the k to h line equals the ratio of collision strengths, as the escape probability of photons is unity. As the  \ion{Mg}{2} k and h lines share the same ionization state and originate from a transition to a shared lower level, and given that the statistical weight ($\omega_{i}$) is the same in both cases, the line intensity ratio is simply the ratio of oscillator strengths ($f_{ij}$). Consequently, this ratio is expected to be 2:1 in optically thin conditions and lower when the medium is optically thick \citep{kerr15,levens19}.

@@

%%--------------------------------------------------
\begin{figure*}[ht!]
    \centering
\includegraphics[trim={7cm 5cm 7.7cm 6cm},clip,width=\textwidth]{Figures/Flare_M_Nov04_2015_2.eps}
\caption{The M3.7 flare observed on November 4th, 2015. Panel a: GOES flux plot in 0.5{--}4~{\AA} (blue) and 1.0{--}8.0~{\AA} (red). Panel b: AIA 1600~{\AA} image of the flaring region. Arrows locate the primary ribbons. Panel c: LOS magnetic flux density map obtained from HMI near the peak of the flare. The over-plotted white (black) boxes in panel b (c) represent the IRIS SJI FOV. The over-plotted white dot-dashed (magenta dashed) box in panel b (c) shows the IRIS raster FOV.}\label{flare1}
\end{figure*}
%%--------------------------------------------------

@@

%%%--------------
\begin{figure*}[ht!]
    \begin{center}
    \includegraphics[width=\textwidth]{Figures/nov_flare_304_aa_evolv.eps}
    \end{center}
    \caption{Sequence of AIA 304~{\AA}~images for the Nov 4th, 2015 flare. The white dotted box shows the portion of the IRIS raster FOV which scanned the ribbons. In panels b (c) F1 and F2 are the filament material, which move away from each other as the flare progresses. R1 and R2 in b (c,d) show the primary ribbons which move through the IRIS raster FOV. An animation of this image sequence is available in \cite{roy24}.}
    \label{flare_m_ev}
\end{figure*}
%%%--------------

@@

%%%--------------
\begin{figure*}[ht!]
    \begin{center}
        \includegraphics[trim={4.5cm 5.5cm 6cm 4cm},clip,width=\textwidth]{Figures/nov_flare_aia_waves_2.pdf}
    \end{center}
    \caption{The Nov 4th, 2015 flare in the six Coronal channels of the SDO/AIA at the soft X-ray peak (panels (a)-(f)) and during the decline phase (panels (g)-(l)) of the flare.}
    \label{flare_m_aia}
\end{figure*}
%%%--------------

@@

%%%%-------------------------
\begin{figure}[ht!]
\centering  
\vspace{10cm}
\includegraphics[trim={7cm 5cm 6cm 15cm},width=0.7\textwidth]{Figures/m_flare_iris.pdf}
\caption{Obtained images in  \ion{Mg}{2} h (panel a) and k (panel b) and corresponding co-aligned and artificially rastered HMI LOS magnetic field map (panel c) for the M3.7 flare observed on November 4th, 2015. The magenta and lime green contours on panel c show the contours of  \ion{Mg}{2}~h (panel a) and k (panel b) intensity.} \label{fig:aligned_raster}
\end{figure}
%%%%-------------------------

@@

%%%%%%%%%%%
\begin{figure*}[ht!]
    \centering
    \includegraphics[trim={0cm 3cm 0.5cm 3cm},clip,width=\textwidth]{Figures/pix_fit.pdf}
    \caption{Fit of the  \ion{Mg}{2} window for the pixel marked in northern ribbon (southern ribbon) in Fig.~\ref{fig:aligned_raster} in top panel (bottom panel) from various times during the evolution of the flare. }
    \label{fig:pix_fit_ribbon}
\end{figure*}
%%%%%%%%%%%

%%%%-------------------------
\begin{figure}[ht!]
\centering  
\includegraphics[trim={1.5cm 4cm 0.5cm 4cm},clip,width=0.8\textwidth]{Figures/m_flare_iris_pt2.pdf}
\caption{The 1{--}8~{\AA} GOES light curve of the flare over-plotted with the time variation of  \ion{Mg}{2} k/h line intensity ratio for the northern (black) and southern (green) asterisks marked in panels a, b \& c.}
\label{fig:aligned_iris_ratio}
\end{figure}
%%%%-------------------------

@@

%%----------------------------------------------------------------------------
\begin{figure}[ht!]
    \centering
    \includegraphics[trim={8cm 1cm 2cm 0.2cm},clip,width=0.9\textwidth]{Figures/Flare-m-optical-depth-2.jpeg}
    \caption{ \ion{Mg}{2} k to h line intensity ratio for various magnetic flux density bins at various time steps during the evolution of the flare.}
    \label{fig:optical_depth_m}
\end{figure}
%%----------------------------------------------------------------------------

@@

%%%----------------------------------------------
\begin{figure*}[ht!]
    \centering
    \includegraphics[trim={3cm 3cm 2cm 4cm},clip,width=\textwidth]{Figures/Nov-11-2015-optical-dep-ev-5.pdf}
    \caption{Time evolution of the  \ion{Mg}{2} k to h line intensity ratio obtained from averaged spectrum over the corresponding magnetic flux bin as labeled for the Nov 4, 2015 flare. For better visibility, the 20.9{--}184.4 G and 348.9{--}513.3 G points are offset by 30s and -30s respectively. Over-plotted red solid line displays the 1{--}8~{\AA} GOES X-ray light curve.}
    \label{fig:optical_dep_ev_m}
\end{figure*}
%%%----------------------------------------------

@@

%%--------------------------------------------------
\begin{figure}[ht!]
    \centering
\hspace*{-.5in}
\includegraphics[width=\textwidth,trim={7cm 5cm 7.7cm 6cm},clip]{Figures/Flare_X_oct22_2014_2.eps}
\caption{The X class flare observed on October 22, 2014. Panel a: GOES flux plot in 0.5{--}4~{\AA} (blue) and 1.0{--}8.0~{\AA} (red). Panel b: AIA 1600~{\AA} image taken at the peak of the flare. Arrows locate the primary and secondary ribbons. Panel c: LOS magnetic flux density map obtained from HMI at the peak of the flare. The over-plotted white (black) box in panel b (c) represents the IRIS SJI FOV, and white dot-dashed (dashed) box in panels b (c) represents the IRIS Raster FOV.}\label{flare2}
\end{figure}
%%--------------------------------------------------

@@

%%--------------------------------------------------
\begin{figure}[ht!]
    \centering
%    \hspace{2in}
    \includegraphics[trim={6cm 3cm 6cm 3cm},clip,width=0.45\textwidth]{Figures/contour_paper_plot_oct28.pdf}
    \includegraphics[trim={3cm 2cm 2cm 3cm},clip,width=0.45\textwidth]{Figures/Oct-22-2014-optical-dep-ev-2.pdf}
    \caption{Obtained intensity maps in  \ion{Mg}{2}~k (panel a) and h (panel b) lines at the peak of the flare. The rastrered HMI LOS magnetic field density map is shown in panel c. The black and lime green contours on panel c show the contours of  \ion{Mg}{2}~k (panel a) and h (panel b) intensity.}
\label{fig:align_raster_flare2}
\end{figure}
%%--------------------------------------------------

@@

%%--------------------------------------------------
\begin{figure*}[ht!]
    \centering
\hspace*{-.6in}
\includegraphics[width=1.12\textwidth,trim={7cm 5cm 7.7cm 6cm},clip]{Figures/Flare_C_feb03_2015_2.eps}
%\includegraphics[width=.7\textwidth]{figures/Feb_03_2015/Feb-04-2015-optical-dep-ev-2.eps}
\caption{The C class flare observed on February 3rd, 2015. Panel a: GOES flux plot in 0.5{--}4~{\AA} (blue) and 1.0{--}8.0~{\AA} (red). Panel b: AIA 1600~{\AA} image of the flaring region. Panel c: LOS magnetic flux density map obtained from HMI near the peak of the flare. The over-plotted white (black) boxes in panel b (c) represents the IRIS SJI FOV. The over-plotted white dot-dashed (magenta dashed) box in panel b (c) show the IRIS raster FOV.}\label{flare3}
\end{figure*}
%%--------------------------------------------------

%%--------------------------------------------------
\begin{figure}[ht!]
    \centering
%\hspace*{-1.in}
\includegraphics[width=0.5\textwidth,trim={4cm 5cm 4cm 4cm},clip]{Figures/contour_paper_plot_feb-03-2015.pdf}
\caption{Obtained intensity maps in  \ion{Mg}{2}~k (panel a) and h (panel b) lines at the peak of the flare. The rastrered HMI LOS magnetic field density map is shown in panel c. The black and lime green contours on panel c show the contours of  \ion{Mg}{2}~k (panel a) and h (panel b) intensity.}
\label{fig:align_raster_flare3}
\end{figure}
%%--------------------------------------------------

%%--------------------------------------------------
\begin{figure}[ht!]
    \centering
    \includegraphics[trim={2cm 3cm 2cm 3cm},clip,width=0.8\textwidth]{Figures/Feb-04-2015-optical-dep-ev-2.pdf}
    \caption{Time evolution of the  \ion{Mg}{2} k to h line intensity ratio obtained from averaged spectrum over the corresponding magnetic flux bin as labeled for the Feb 3, 2015 flare. Over-plotted red solid line displays the 1{--}8~{\AA} GOES X-ray light curve.}
    \label{fig:optical_dep_ev_c}
\end{figure}
%%--------------------------------------------------

@@

%%--------------------------------------------------
\section{Outlook}
%%--------------------------------------------------

The  \ion{Mg}{2}~k to h line intensity ratios serve as a valuable diagnostic for assessing the opacity of the local plasma and can provide insights into changes in the local plasma environment within the solar chromosphere during flares. In this study, we investigated the temporal variation of  \ion{Mg}{2}k to h line intensity ratios during the evolution of three flares classified as X-class, M-class, and C-class. We also examined the variation of intensity ratios across different magnetic flux density bins, utilizing observations from IRIS and HMI. Co-alignment of IRIS and HMI observations was achieved using 1600{\AA} images recorded by AIA.

>>>>

%%---------------------------------------------------------------
\begin{equation}
\mathrm{\Omega_{ij}=\frac{8\pi}{\sqrt{3}}~\frac{I_{H}}{\Delta \epsilon_{ij}}g\omega_{i}~f_{ij}}
\end{equation}
%%---------------------------------------------------------------

Here, $I_{H}$ denotes the ionization energy of hydrogen, $\Delta \epsilon_{ij}$ represents the threshold energy for the transition, $g$ is the Gaunt factor, $\omega_{i}$ is the statistical weight of the level, and $f_{ij}$ stands for the oscillator strength. In optically thin conditions, the intensity ratio of the k to h line equals the ratio of collision strengths, as the escape probability of photons is unity. As the  \ion{Mg}{2} k and h lines share the same ionization state and originate from a transition to a shared lower level, and given that the statistical weight ($\omega_{i}$) is the same in both cases, the line intensity ratio is simply the ratio of oscillator strengths ($f_{ij}$). Consequently, this ratio is expected to be 2:1 in optically thin conditions \citep{kerr15,levens19}.{\bf For further details on the ratio of the oscilator strengths please refer to \S\ref{sec:a1}.}

{\bf Under optically thick conditions the radiation is absorbed and re-emitted more effectively by the medium. As a result the radiation field in the medium becomes "thermalized", meaning the population of the energy levels are dictated by the Boltzmann distribution, rather than the specific radiative process or the specific line strengths. As a result the intensity of the lines are no longer individually dependent on their Einstein coefficients or the detailed radiative processes but are instead controlled by the overall temperature and the combined behavior of the radiation field in an optically thick medium. This leads to a situation where both the upper levels for the \ion{Mg}{2}~k~and~h lines are very similar, as they are formed at relatively similar heights. So the intensity ratio is closer to 1:1 \citep{kerr15, levens19}. This ratio has been used before to comment on the opacity of the local plasma environment. It has been demonstrated from previous solar observations, that the \ion{Mg}{2} lines exhibit no central reversal $\sim$ 6\arcsec~above the limb, and $\sim$ 12\arcsec~above the limb the line intensity ratio is 2:1, implying that the lines become optically thin \citep{doschek_n_feldman_77, feldman_n_doscheck_77}. on the contrary, \cite{kneer81} showed from sunspot spectra, although the \ion{Mg}{2} lines show no central reversal, the ratio was still lower 2:1, implying optically thick conditions.
}

@@

%%--------------------------------------------------
\begin{figure*}[ht!]
    \centering
\includegraphics[trim={0.4cm 7cm 2.5cm 6.5cm},clip,width=\linewidth]{Figures/Flare_M_Nov04_2015_2.pdf}
\caption[The observations of the M3.7 flare on November 4th, 2015]{The M3.7 flare observed on November 4th, 2015. Panel a: GOES flux plot in 0.5{--}4~{\AA} (blue) and 1.0{--}8.0~{\AA} (red). Panel b: AIA 1600~{\AA} image of the flaring region. Arrows locate the primary ribbons. Panel c: LOS magnetic flux density map obtained from HMI near the peak of the flare. The over-plotted white (black) boxes in panel b (c) represent the IRIS SJI FOV. The over-plotted white dot-dashed (magenta dashed) box in panel b (c) shows the IRIS raster FOV.}\label{flare1}
\end{figure*}
%%--------------------------------------------------

@@

%%%--------------
\begin{figure*}[ht!]
    \begin{center}
    \includegraphics[trim={1cm 1.2cm 1cm 2.3cm}, clip, width=\textwidth]{Figures/nov_flare_304_aa_evolv.eps}
    \end{center}
    \caption[Sequence of AIA 304 {\AA} images for the Nov 4th, 2015 flare.]{Sequence of AIA 304~{\AA}~images for the Nov 4th, 2015 flare. The white dotted box shows the portion of the IRIS raster FOV which scanned the ribbons. In panels b (c) F1 and F2 are the filament material, which move away from each other as the flare progresses. R1 and R2 in b (c,d) show the primary ribbons which move through the IRIS raster FOV. An animation of this image sequence is available in \cite{roy24}.}
    \label{flare_m_ev}
\end{figure*}
%%%--------------

@@

%%%--------------
\begin{figure*}[ht!]
    \begin{center}
        \includegraphics[trim={4.7cm 5.5cm 6.5cm 4.2cm},clip,width=\textwidth]{Figures/nov_flare_aia_waves_2.pdf}
    \end{center}
    \caption[The six coronal channel observations from SDO/AIA for the Nov 4th, 2015 flare.]{The Nov 4th, 2015 flare in the six Coronal channels of the SDO/AIA at the soft X-ray peak (panels (a)-(f)) and during the decline phase (panels (g)-(l)) of the flare.}
    \label{flare_m_aia}
\end{figure*}
%%%--------------

@@

%%%%-------------------------
\begin{figure}[ht!]
\centering  
\vspace{10cm}
\includegraphics[trim={7cm 5cm 6cm 15cm},width=0.7\textwidth]{Figures/m_flare_iris.pdf}
\caption[\ion{Mg}{2}~h and k line intensity observation with artificially rastered HMI LOS magnetic field for the November 4th, 2015 flare.]{Obtained images in \ion{Mg}{2}~h (panel a) and k (panel b) and corresponding co-aligned and artificially rastered HMI LOS magnetic field map (panel c) for the M3.7 flare observed on November 4th, 2015. The magenta and lime green contours on panel c show the contours of  \ion{Mg}{2}~h (panel a) and k (panel b) intensity.} \label{fig:aligned_raster}
\end{figure}
%%%%-------------------------

@@

%%%%%%%%%%%
\begin{figure*}[ht!]
    \centering
    \includegraphics[trim={0cm 3cm 0.5cm 3cm},clip,width=\textwidth]{Figures/pix_fit.pdf}
    \caption[Fit of the \ion{Mg}{2}~window for a pixel.]{Fit of the \ion{Mg}{2} window for the pixel marked in northern ribbon (southern ribbon) in Fig.~\ref{fig:aligned_raster} in top panel (bottom panel) from various times during the evolution of the flare. }
    \label{fig:pix_fit_ribbon}
\end{figure*}
%%%%%%%%%%%

%%%%-------------------------
\begin{figure}[ht!]
\centering  
\includegraphics[trim={1.5cm 4cm 0.5cm 4cm},clip,width=0.8\textwidth]{Figures/m_flare_iris_pt2.pdf}
\caption[The 1{--}8~{\AA} GOES light curve of the flare over-plotted with the time variation of \ion{Mg}{2}~k/h line intensity ratio.]{The 1{--}8~{\AA} GOES light curve of the flare over-plotted with the time variation of  \ion{Mg}{2} k/h line intensity ratio for the northern (black) and southern (green) asterisks marked in panels a, b \& c.}
\label{fig:aligned_iris_ratio}
\end{figure}
%%%%-------------------------

@@

%%----------------------------------------------------------------------------
\begin{figure}[ht!]
    \centering
    \includegraphics[trim={8cm 1cm 2cm 0.2cm},clip,width=0.9\textwidth]{Figures/Flare-m-optical-depth-2.jpeg}
    \caption{\ion{Mg}{2}~k to h line intensity ratio for various magnetic flux density bins at various time steps during the evolution of the flare.}
    \label{fig:optical_depth_m}
\end{figure}
%%----------------------------------------------------------------------------

@@

%%%----------------------------------------------
\begin{figure*}[ht!]
    \centering
    \includegraphics[trim={3cm 3cm 2cm 4cm},clip,width=\textwidth]{Figures/Nov-11-2015-optical-dep-ev-5.pdf}
    \caption[Time evolution of the \ion{Mg}{2}~k to h line intensity ratio obtained from averaged spectrum over the corresponding magnetic flux bin.]{Time evolution of the \ion{Mg}{2}~k to h line intensity ratio obtained from averaged spectrum over the corresponding magnetic flux bin as labeled for the Nov 4, 2015 flare. For better visibility, the 20.9{--}184.4 G and 348.9{--}513.3 G points are offset by 30s and -30s respectively. Over-plotted red solid line displays the 1{--}8~{\AA} GOES X-ray light curve.}
    \label{fig:optical_dep_ev_m}
\end{figure*}
%%%----------------------------------------------

@@

%%--------------------------------------------------
\begin{figure}[ht!]
    \centering
\includegraphics[width=\textwidth,trim={0.5cm 7cm 3cm 4.5cm},clip]{Figures/Flare_X_oct22_2013.pdf}
\caption[The observations of the X flare on November 4th, 2015]{The X class flare observed on October 22, 2014. Panel a: GOES flux plot in 0.5{--}4~{\AA} (blue) and 1.0{--}8.0~{\AA} (red). Panel b: AIA 1600~{\AA} image taken at the peak of the flare. Arrows locate the primary and secondary ribbons. Panel c: LOS magnetic flux density map obtained from HMI at the peak of the flare. The over-plotted white (black) box in panel b (c) represents the IRIS SJI FOV, and white dot-dashed (dashed) box in panels b (c) represents the IRIS Raster FOV.}\label{flare2}
\end{figure}
%%--------------------------------------------------

@@

%%--------------------------------------------------
\begin{figure}[ht!]
    \centering
%    \hspace{2in}
    \includegraphics[trim={0.1cm 0.2cm 8cm 0.2cm},clip,width=0.5\textwidth]{Figures/contour_paper_plot_oct28.pdf}
    \includegraphics[trim={0.1cm 1cm 0.1cm 1cm},clip,width=0.7\textwidth]{Figures/Oct-22-2014-optical-dep-ev-2.pdf}
    \caption[Top panel: \ion{Mg}{2}~h and k line intensity observation with artificially rastered HMI LoS magnetic field for the October 22, 2014 flare.]{Obtained intensity maps in  \ion{Mg}{2}~k (panel a) and h (panel b) lines at the peak of the flare. The rastrered HMI LoS magnetic field density map is shown in panel c. The black and lime green contours on panel c show the contours of  \ion{Mg}{2}~k (panel a) and h (panel b) intensity. Bottom panel: Time evolution of the \ion{Mg}{2}~k to h line intensity ratio obtained from averaged spectrum over the corresponding magnetic flux bin as labeled for the Oct 22, 2014 flare. The over-plotted red solid line displays the 1{--}8~{\AA} GOES X-ray light curve.}
\label{fig:align_raster_flare2}
\end{figure}
%%--------------------------------------------------

@@

%%--------------------------------------------------
\begin{figure*}[ht!]
    \centering
\includegraphics[width=\textwidth,trim={0cm 4.5cm 3cm 4cm},clip]{Figures/Flare_C_feb03_2015.pdf}
%\includegraphics[width=.7\textwidth]{figures/Feb_03_2015/Feb-04-2015-optical-dep-ev-2.eps}
\caption[The observations of the X flare on February 3rd, 2015.]{The C class flare observed on February 3rd, 2015. Panel a: GOES flux plot in 0.5{--}4~{\AA} (blue) and 1.0{--}8.0~{\AA} (red). Panel b: AIA 1600~{\AA} image of the flaring region. Panel c: LOS magnetic flux density map obtained from HMI near the peak of the flare. The over-plotted white (black) boxes in panel b (c) represents the IRIS SJI FOV. The over-plotted white dot-dashed (magenta dashed) box in panel b (c) show the IRIS raster FOV.}\label{flare3}
\end{figure*}
%%--------------------------------------------------

%%--------------------------------------------------
\begin{figure}[ht!]
    \centering
%\hspace*{-1.in}
\includegraphics[width=0.7\textwidth,trim={4cm 5cm 4cm 4cm},clip]{Figures/contour_paper_plot_feb-03-2015.pdf}
\caption[\ion{Mg}{2}~h and k line intensity observation with artificially rastered HMI LOS magnetic field for the February 3rd, 2015 flare.]{Obtained intensity maps in  \ion{Mg}{2}~k (panel a) and h (panel b) lines at the peak of the flare. The rastrered HMI LOS magnetic field density map is shown in panel c. The black and lime green contours on panel c show the contours of  \ion{Mg}{2}~k (panel a) and h (panel b) intensity.}
\label{fig:align_raster_flare3}
\end{figure}
%%--------------------------------------------------

%%--------------------------------------------------
\begin{figure}[ht!]
    \centering
    \includegraphics[trim={2cm 3cm 2cm 3cm},clip,width=0.8\textwidth]{Figures/Feb-04-2015-optical-dep-ev-2.pdf}
    \caption[Time evolution of the \ion{Mg}{2}~k to h line intensity ratio obtained from averaged spectrum over the corresponding magnetic flux bin for the February 3rd, 2015 flare.]{Time evolution of the \ion{Mg}{2}~k to h line intensity ratio obtained from averaged spectrum over the corresponding magnetic flux bin as labeled for the Feb 3, 2015 flare. The over-plotted red solid line displays the 1{--}8~{\AA} GOES X-ray light curve.}
    \label{fig:optical_dep_ev_c}
\end{figure}
%%--------------------------------------------------

@@

%%--------------------------------------------------
\section{Summary and Discussion}
%%--------------------------------------------------

The  \ion{Mg}{2}~k to h line intensity ratios serve as a valuable diagnostic for assessing the opacity of the local plasma and can provide insights into changes in the local plasma environment within the solar chromosphere during flares. In this study, we investigated the temporal variation of  \ion{Mg}{2}k to h line intensity ratios during the evolution of three X-class, M-class, and C-class flares. We also examined the variation of intensity ratios across different magnetic flux density bins, utilizing observations from IRIS and HMI. Co-alignment of IRIS and HMI observations was achieved using 1600{\AA} images recorded by AIA.
